Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease

### 2023-05-13 doc/manuscript/translocation_pnas.tex
Convert manuscript to PNAS after small edits

% Please add a significance statement to explain the relevance of your work
\significancestatement{Understanding how species persist despite accelerating global change is
critical for the conservation of biodiversity. One possibility is
evolutionary modification that restores positive growth to declining
populations (``evolutionary rescue''). Individuals from these rescued
populations could also be used to reestablish exinct populations. We
evaluated both possibilities in a once-common frog species driven to the
brink of extinction by a disease that has devastated global amphibian
biodiversity and for which few options exist to reverse its impacts. We
found that frogs have evolved increased resistance to infection,
facilitating population recovery. In addition, introduction of frogs
from these populations allowed reestablishment despite the ongoing
presence of disease. Such reintroductions could play an important role
in biodiversity conservation amidst unprecedented global change.}

@@

\begin{abstract}
Escalating anthropogenic change is devastating global biodiversity. To
avoid extinction, affected populations will often need to adapt to the
altered environment. This process of rapid evolutionary change that
increases the frequency of adaptive alleles and thereby restores
positive population growth is termed ``evolutionary rescue''. In
addition to allowing recovery of extant populations, individuals from
rescued populations could also be used to reestablish extirpated
populations. We evaluated this scenario using the once-common mountain
yellow-legged (MYL) frog, which is threatened with extinction by the
human-facilitated emergence of a lethal fungal pathogen
(\emph{Batrachochytrium dendrobatidis}; ``Bd''). Although most MYL frog
populations are extirpated following disease outbreaks, some persist and
eventually begin to recover. We show that MYL frogs have undergone
substantial evolutionary change following disease outbreaks, and that
these changes are associated with increased resistance/tolerance to Bd
infection in naturally recovering populations. Large-scale
reintroduction of frogs from these populations resulted in the
establishment of reproducing populations at most recipient sites despite
ongoing disease. In addition, results from viability modeling suggest
that established populations have a low probability of extinction over
50 years. Collectively, these results provide one of the first examples
of how evolutionary rescue and the reintroduction of resistant/tolerant
individuals can allow the landscape-scale recovery of disease-impacted
species. This example has potentially broad implications for the many
taxa worldwide that are threatened with extinction by anthropogenic
stressors.
\end{abstract}

@@

\dropcap{H}uman activities are intensifying threats to global biodiversity
\citep{ceballos2015}, with important implications for ecosystem
resilience and human welfare \citep{naeem2009}. To persist, affected
taxa will often need to adapt to their altered environment or face
decline and extinction. ``Evolutionary rescue'' describes rapid
evolutionary change that increases the frequency of adaptive alleles and
thereby restores positive population growth \citep{carlson2014}. In
addition to preventing extinction of declining populations, in theory,
this process can also result in the seeding of nearby empty habitats and
allow population reestablishment and recovery. Therefore, understanding
the mechanisms supporting population persistence, including evolutionary
rescue, is critical for the conservation of the ever-growing number of
threatened taxa.

One consequence of human alteration of the biosphere is an increase in
emerging infectious diseases \citep{jones2008, fisher2012}. Such
diseases pose a severe threat to wildlife populations
\citep{daszak2000}, and have caused dramatic declines and extinctions in
a wide range of taxa
\citep{hewson2014, samuel2015, scheele2019, cunningham2021}. Amid this
global loss of biodiverity, some populations previously decimated by
disease have subsequently recovered to varying degrees
\citep{newell2013, voyles2018, knapp2016}. Although we know relatively
little about the underlying mechanisms
\citep{brannelly2021, russell2020}, host evolution is one possibility
\citep{carlson2014, searle2020}. This process is known from a diversity
of wildlife populations
\citep{savage2016, epstein2016, gignoux-wolfsohn2021, holland2022}, and
often results in increased host resistance and/or tolerance (see
\citep{medzhitov2012} for definitions). However, linking these
evolutionary responses to population recoveries - as necessary for
evolutionary rescue - is difficult and few clear examples exist. In
addition to promoting recovery of disease-impacted populations, the
resistant/tolerant individuals in such populations could potentially be
used to reestablish populations that were extirpated following disease
emergence \citep[e.g.,][]{mendelson2019, brannelly2021}. To date, the
use of resistant/tolerant individuals for species recovery is rare
\citep[but see][]{joseph2018}, and studies that evaluate its potential
utility are critically needed.

The mountain yellow-legged (MYL) frog, composed of the sister species
\emph{Rana muscosa} and \emph{Rana sierrae} \citep{vredenburg2007}, is
emblematic of the unprecedented global declines of amphibians caused by
the recently-emerged amphibian chytrid fungus
\citep[\emph{Batrachochytrium dendrobatidis,}][]{scheele2019}. Once the
most common amphibian in the high elevation portion of California's
Sierra Nevada mountains \citep[USA,][]{grinnell1924}, this frog has
disappeared from more than 90\% of its historical range during the past
century \citep{vredenburg2007}. Due to the severity of its decline and
the increasing probability of extinction, both species are now listed as
``endangered'' under the U.S. Endangered Species Act. In the Sierra
Nevada, this decline was initiated by the introduction of non-native
trout into the extensive historically-fishless region
\citep{bradford1989, knapp2000}, and the arrival of \emph{B.
dendrobatidis} (Bd) in the 1960s and its subsequent spread
\citep{vredenburg2019} caused additional widespread population
extirpations \citep{vredenburg2010, rachowicz2006}. These Bd-caused
declines are fundamentally different from the fish-caused declines
because removal of fish populations is feasible and results in the rapid
recovery of frogs \citep{knapp2007, vredenburg2004}. In contrast, Bd
appears to persist in habitats even in the absence of amphibian hosts
\citep{walker2007}, and therefore represents a long-term alteration of
invaded ecosystems that amphibians will need to overcome to reestablish
populations.

@@

In the current study, we combine results from genomic analyses, a
15-year landscape-scale frog reintroduction effort, and detailed
population models to provide an unprecedented example of how evolution
in response to disease allows the successful reestablishment of
extirpated MYL frog populations. Specifically, we used exome capture
methods to show that MYL frogs have undergone substantial evolutionary
change associated with Bd exposure and that may provide a genetic basis
for the observed increased resistance/tolerance to Bd infection observed
in naturally recovering populations \citep{knapp2016}. We then describe
the reintroduction of these resistant/tolerant frogs to 24 sites and the
dynamics and drivers of population establishment using results from
long-term capture-mark-recapture (CMR) surveys. These data are also used
to parameterize a population model that extends our inferences of
population establishment well beyond the temporal extent of the CMR
data. Collectively, these results provide one of the first examples of
how evolutionary rescue and the reintroduction of resistant/tolerant
animals can allow the large-scale recovery of disease-impacted species.

@@

To determine whether frog populations show an evolutionary response to
Bd, we compared genomes of frogs in populations with different histories
of Bd exposure. Specifically, we compared frog genomes sampled in four
populations that have not yet experienced a Bd-caused epizootic
(``naive'') \citep{zhou2015} versus in five populations that experienced
a Bd epizootic during the past several decades and have since recovered
to varying degrees (``recovering''; Figure~\ref{fig-allelefrequencies})
\citep{knapp2016, vredenburg2010}. These recovering populations persist
in an enzootic state \citep{briggs2010}, characterized by high Bd
prevalence and, in adults, moderate Bd infection intensities (``load'')
that are typically well below the level expected to cause mortality
\citep{vredenburg2010}.

We conducted a principal component analysis (PCA) of the genomic data to
describe the association between sampled populations, and then used two
complementary approaches to identify regions of the genome under
selection. First, we used a multivariate linear mixed model to evaluate
associations between population type (i.e., naive versus recovering) and
specific variants, including single nucleotide polymorphisms (SNPs) and
insertions/deletions (INDELS). Second, we used a splined window analysis
to identify larger genomic regions showing differences between
population types in \emph{F\textsubscript{ST}} and nucleotide diversity
(\(\pi_{diff}\) = \(\pi_{naive}\) - \(\pi_{recovering}\)).

Individual frogs clustered into 3 separate groups in principal component
space (Figure S1A), and clusters reflected
the strong signature of isolation by distance that is characteristic of
MYL frogs \citep{rothstein2020, poorten2017}. Importantly, each cluster
contained at least one population from both the naive and recovering
groups, allowing us to distinguish between (i) allelic associations of
individuals sampled in naive versus recovering populations, and (ii)
allelic associations resulting from population structure and genetic
drift.

Results from the SNP and splined window analyses show that recovering
populations have signatures of strong selection on multiple regions of
the genome. The SNP analysis identified 11 outlier SNPs from 7 distinct
genes across 4 contigs (Figure S1B,C, SI
Dataset 1). One of the 7 identified genes did not have an associated
annotation. Frequency differences between the naive and recovering
populations ranged from 41\% to 86\%. Most of these SNPs showed
frequency differences in only a subset of the sampled populations
(Figure~\ref{fig-allelefrequencies} A, B), but the outlier SNP in the
RIN3 gene showed consistent differences in frequencies across all
populations (Figure~\ref{fig-allelefrequencies} C). This is suggestive
of strong - and parallel - selection at this locus associated with
exposure of frogs to Bd across multiple populations.

The splined window analysis identified 58 outlier regions for
\emph{F\textsubscript{ST}} and 33 outlier regions for \(\pi_{diff}\)
(Figure~\ref{fig-spline-manhattan} A, B). Of these, 9 regions were
outliers for both metrics and 2 of these shared regions also contained
outlier SNPs. A total of 33 annotated genes were found in the 9 shared
regions (SI Dataset 2). The largest \(\pi_{diff}\), indicative of
directional selection, occurred in a 163kb region on Contig19, 12.9Mb
upstream of the RIN3 outlier SNP (Figure~\ref{fig-spline-manhattan} C).
This region contains approximately 500 SNPs and one annotated gene
called ``interferon-induced very large GTPase 1-like'' (GVINP1).
Additionally, a shared outlier region on Contig1 contained two
complement factor genes (C6 and C7). Interestingly, this region had a
low \(\pi_{diff}\) value, consistent with balancing selection. Finally,
one shared outlier region on Contig8 contained one outlier SNP (TCF19)
and was within 360kb of another outlier SNP (VARS)
(Figure~\ref{fig-spline-manhattan} D, Figure S2).
This region (854kb from the beginning of the outlier window to the VARS
SNP) contained a total of 8 annotated genes. In \emph{Xenopus}, five of
these genes occur in the extended major histocompatibility complex (MHC)
Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the
extended MHC Class III region (HSP70, LSM2, VARS) \citep{ohta2006}.
Therefore, this region under selection is part of the extended MHC Class
I and III complex and shows synteny with other amphibian genomes.

@@

To determine whether Bd load had a negative effect on adult survival, as
would be expected if frogs were highly susceptible to Bd infection, we
developed linear and multilevel models. The best model of 1-year frog
survival identified several important predictors, but Bd load at the
time of translocation was not among them (bd\_load;
Figure S6). Important predictors included
winter severity in the year following translocation (snow\_t1), site
elevation (elevation), and donor population (donor\_72996;
Figure~\ref{fig-cond-effects}). Males had somewhat higher survival than
females, but this effect was only marginally important (sex\_male;
Figure~\ref{fig-cond-effects}, Figure S6).
Collectively, these results are consistent with frogs in recovering
populations having evolved sufficient resistance/tolerance to minimize
the adverse effects of Bd infection. In addition, the moderate Bd loads
that characterized frogs following translocation
(Figure S5) and relatively high survival of
frogs in numerous translocated cohorts
(Figure~\ref{fig-translocation-survival}) suggests that frogs
translocated from recovering populations can maintain the benefits of
resistance/tolerance in non-natal habitats. Finally, in 3 locations
where longer CMR time series allowed us to assess the survival of new
adults recruited to the population, naturally recruited adults had
equivalent or higher survival probabilities than the originally
translocated adults (Figure S7). Therefore,
frog resistance in non-natal habitats is maintained across generations.

@@

Even when incorporating demographic stochasticity and environmental
stochasticity in the probability of juvenile recruitment to the adult
class \(\omega\) (the transition that we expect to be the most subject
to environmental variability in the presence of Bd), populations with
high adult survival are likely to persist over a 50 year time horizon.
Our model predicted that, following a single introduction of 40 adult
individuals into a population and using the baseline value for
\(\omega\) of 0.3, the six populations with the highest adult survival
probabilities had 50-year extinction probabilities of 0 to 0.32;
Figure~\ref{fig-viability} B). This indicates strong potential for
long-term persistence in the presence of Bd and environmental
variability in recruitment (Figure~\ref{fig-viability} B). When we
decreased \(\omega\) from 0.3 to a conservative estimate of 0.15,
50-year extinction probability remained less than 0.5 for three
populations and was 0.79 to 0.94 for three additional populations
(Figure~\ref{fig-viability} B). In contrast, for the six populations
where \(\sigma_{A_R} < 0.5\), extinction probability over 50 years was
always predicted to be 1. To test the validity of our model predictions,
we demonstrated that our stochastic model could describe the general
recovery trajectory of our translocated population with the longest
survey history (Figure~\ref{fig-viability} C; population 70550, surveyed
for 16 years).

@@

Results from our genomic analyses strongly suggest that natural
selection for adaptive alleles and resulting increased
resistance/tolerance is at least partially responsible for the reduced
susceptibility of frogs in recovering populations. We identified
specific alleles and genomic regions showing signatures of selection
between adjacent Bd-naive and recovering MYL frog populations,
consistent with strong selection following Bd exposure. Importantly,
some regions under selection are associated with cellular and
immunological mechanisms known to contribute to disease resistance,
including in amphibians \citep{zamudio2020}. For example, the MHC plays
an important role in immunity. In our study, we identified a region that
shows evidence of selection in recovering populations and contains eight
genes associated with either the MHC Class I or Class III regions. These
results corroborate numerous previous studies linking MHC genes to
amphibian resistance/tolerance against Bd
\citep[e.g.,][]{savage2011, bataille2015}. Similarly, the region with
the strongest indication of directional selection (as measured by
\(\pi_{diff}\)) contains the gene GVINP1. Several previous studies have
found this gene to be differentially expressed during Bd infection
\citep[e.g.,][]{grogan2018, ellison2014}. This gene is also strongly
linked to disease in salmon, explaining a notable 20\% of the resistance
phenotype \citep{robledo2020, robledo2018}. Finally, the RIN3 gene
showed a consistent pattern of allele frequency differences across all
nine of the frog populations sampled in this study, indicating
consistent selection in populations distributed across a wide geographic
area. This gene is associated with inflammation response and in
\emph{Xenopus} is highly expressed in wound stumps following amputation
of tadpole tails \citep{fukazawa2009}. In the SI, we provide additional
details on these and other genes identified in our study as under
selection.

@@

To visualize the genomic relationship of our populations we conducted a
principal components analysis (PCA) using the glPCA function in the
adegenet R package \citep{jombart2008}. To detect regions of the genome
under selection between naive and recovering populations, we used two
approches: (1) a multivariate linear mixed model to evaluate individual
variants (SNPs and INDELs), and (2) a splined window analysis to
evaluate larger genomic regions. For the variant analysis, we first used
a stringent data filter to include only variants with \textless{} 5\%
missing data (missing for no more than 2 individuals), and then
calculated the likelihood ratio statistic for the resulting set of
148,307 high quality variants across 127 contigs using GEMMA. GEMMA
calculates and incorporates a relatedness matrix for input samples,
allowing us to account for relatedness and population structure when
calculating likelihood ratio statistics. We identified ``outlier''
variants using a Bonferroni- corrected significance level of 0.01.
Outliers are variants showing different allele frequencies between naive
versus recovering populations. We visualized the results using a
Manhattan plot and qqplot. We developed a more liberal set of outlier
variants using a Bonferroni- corrected significance level of 0.05 and
used this set solely for the gene ontology (GO) analysis (see below).

@@

We estimated survival of translocated frogs, recruitment of new frogs
into the adult population, and adult population size using a Bayesian
open-population Jolly-Seber mark-recapture model with known additions to
the population (i.e., translocated cohorts), as implemented by the mrmr
package \citep{joseph2019} and using R Statistical Software
\citep[v4.4.4,][]{rsoftware2022}. The model tracks the states of
\emph{M} individuals that comprise a superpopulation made up of real and
pseudo-individuals \citep[see][ for details]{joseph2018}. The possible
states of individuals include ``not recruited'', ``alive'', and
``dead''. The possible observations of individuals include ``detected''
and ``not detected''. We assume that individuals that are in the ``not
recruited'' or ``dead'' states are never detected (i.e., there are no
mistakes in the individual PIT tag records). For all models, we used
mrmr defaults for priors, number of chains (4), and warmup and
post-warmup iterations (2000 for each). We evaluated convergence of the
Markov chain Monte Carlo (MCMC) algorithm using trace plots and
Gelman-Rubin statistics (Rhat). Code used to create CMR datasets and fit
models for each translocated population is available from the following
GitHub repository:
\href{https://github.com/SNARL1/cmr-analysis.}{https://github.com/SNARL1/cmr-analysis}.

@@

After parameterizing our model with CMR-estimated adult frog survival
probabilities and other known vital rates (Table S2), we performed four analyses (described in detail in the SI).
First, we compute the long-run growth rate \(\lambda\) for each of our
12 translocated populations to determine if the populations were
deterministically predicted to growth or decline in the long-run.
Second, we compute the elasticity of \(\lambda\) to model parameters, to
quantify how much changes in adult survival, for example, affected the
long-run growth rate. Third, we included demographic and environmental
stochasticity into our model and simulated the 50-year viability (i.e.,
1 - extinction probability) of populations given an introduction of 40
adult individuals into an unoccupied habitat. Finally, we fit our model
to our longest translocation trajectory to confirm that our model could
reasonably reproduce the observed recovery trajectories of MYL frogs
following reintroductions.}

@@

{\centering \includegraphics[width=0.8\textwidth]{figures/allele_maps.png}

@@

{\centering \includegraphics[width=0.8\textwidth]{figures/splinewindow_manhattan.png}

@@

{\centering \includegraphics[width=0.8\textwidth]{figures/translocation_survival_bysiteid.png}

@@

{\centering \includegraphics[width=0.8\textwidth]{figures/cond_effects_plot.png}

>>>>

% Please add a significance statement to explain the relevance of your work
\significancestatement{Understanding how species persist despite accelerating global change is
critical for the conservation of biodiversity. One possibility is rapid
evolution that restores positive growth to declining populations
(``evolutionary rescue''). Individuals from these rescued populations
could also be used to reestablish exinct populations. We evaluated both
possibilities in a once-common frog species driven to the brink of
extinction by a disease that has devastated global amphibian
biodiversity and for which few options exist to reverse disease impacts.
We found that frogs have evolved increased resistance to infection,
facilitating population recovery. In addition, introduction of frogs
from these populations allowed reestablishment despite the ongoing
presence of disease. Such reintroductions could play an important role
in biodiversity conservation in our rapidly changing world.}

@@

\begin{abstract}
Escalating anthropogenic change is devastating global biodiversity. To
avoid extinction, affected populations will often need to adapt to the
altered environment. The process of rapid evolutionary change that
increases the frequency of adaptive alleles and thereby restores
positive population growth is termed ``evolutionary rescue''. In
addition to allowing recovery of extant populations, individuals from
rescued populations could also be used to reestablish extirpated
populations. We evaluated this scenario in the once-common mountain
yellow-legged (MYL) frog, which is threatened with extinction by the
human-facilitated emergence of a lethal fungal pathogen
(\emph{Batrachochytrium dendrobatidis}; ``Bd''). Although most MYL frog
populations are extirpated following disease outbreaks, some persist and
eventually begin to recover. We show that MYL frogs have undergone
substantial evolutionary change following disease outbreaks, and that
these changes are associated with increased resistance/tolerance to Bd
infection in naturally recovering populations. Large-scale
reintroduction of frogs from these populations resulted in the
establishment of reproducing populations at most recipient sites despite
ongoing disease. In addition, results from viability modeling suggest
that established populations have a low probability of extinction over
50 years. Collectively, these results provide one of the first examples
of how evolutionary rescue and the reintroduction of resistant/tolerant
individuals can allow the landscape-scale recovery of disease-impacted
species. This example has potentially broad implications for the many
taxa worldwide that are threatened with extinction by anthropogenic
stressors.
\end{abstract}

@@

\dropcap{H}uman activities are intensifying threats to global biodiversity
\citep{ceballos2015}, with important implications for ecosystem
resilience and human welfare \citep{naeem2009}. To persist, affected
taxa will often need to adapt to their altered environment or face
decline and extinction. ``Evolutionary rescue'' describes rapid
evolutionary change that increases the frequency of adaptive alleles and
thereby restores positive population growth \citep{carlson2014}. In
addition to preventing extinction of declining populations, this process
could also result in the seeding of nearby empty habitats and allow
population reestablishment and recovery. Although this overall scenario
is potentially of critical importance for the conservation of an
ever-growing number of threatened taxa, descriptions of its operation in
the wild are lacking.

One consequence of human alteration of the biosphere is an increase in
emerging infectious diseases \citep{jones2008, fisher2012}. Such
diseases pose a severe threat to wildlife populations
\citep{daszak2000}, and have caused dramatic declines and extinctions in
a wide range of taxa
\citep{hewson2014, samuel2015, scheele2019, cunningham2021}. Amid this
global loss of biodiverity, some populations decimated by disease have
subsequently recovered to varying extents
\citep{newell2013, voyles2018, knapp2016}. Although we know relatively
little about the mechanisms underlying recovery
\citep{brannelly2021, russell2020}, host evolution is one possibility
\citep{carlson2014, searle2020}. This process is known from a diversity
of wildlife populations
\citep{savage2016, epstein2016, gignoux-wolfsohn2021, holland2022}, and
often results in increased host resistance and/or tolerance (see
\citep{medzhitov2012} for definitions). However, linking these
evolutionary responses to population recoveries is difficult and few
clear examples exist. In addition to promoting recovery of
disease-impacted populations, the resistant/tolerant individuals in such
populations could potentially be used to reestablish populations that
were extirpated following disease emergence
\citep{mendelson2019, brannelly2021}. To date, the use of
resistant/tolerant individuals for species recovery is rare \citep[but
see][]{joseph2018}, and studies that evaluate its potential utility are
critically needed.

The mountain yellow-legged (MYL) frog, composed of the sister species
\emph{Rana muscosa} and \emph{Rana sierrae} \citep{vredenburg2007}, is
emblematic of the unprecedented global declines of amphibians caused by
the recently-emerged amphibian chytrid fungus
\citep[\emph{Batrachochytrium dendrobatidis,}][]{scheele2019}. Once the
most common amphibian in the high elevation portion of California's
Sierra Nevada mountains \citep[USA,][]{grinnell1924}, during the past
century this frog has disappeared from more than 90\% of its historical
range \citep{vredenburg2007}. Due to the severity of its decline and the
increasing probability of extinction, both species are now listed as
``endangered'' under the U.S. Endangered Species Act. In the Sierra
Nevada, this decline was initiated by the introduction of non-native
trout into the extensive historically-fishless region
\citep{bradford1989, knapp2000} starting in the late 1800s. The arrival
of \emph{B. dendrobatidis} (Bd) in the 1960s and its subsequent spread
\citep{vredenburg2019} caused additional large-scale population
extirpations \citep{vredenburg2010, rachowicz2006}. These Bd-caused
declines are fundamentally different from the fish-caused declines
because removal of fish populations is feasible and results in the rapid
recovery of frogs \citep{knapp2007, vredenburg2004}. In contrast, Bd
appears to persist in habitats even in the absence of amphibian hosts
\citep{walker2007}, and therefore represents a long-term alteration of
invaded ecosystems that amphibians will need to overcome to reestablish
populations.

@@

In the current study, we combine results from genomic analyses, a
15-year landscape-scale frog reintroduction effort, and a detailed
population model to provide an unprecedented example of how evolution in
response to disease allows the successful reestablishment of extirpated
MYL frog populations. Specifically, we used exome capture methods to
show that MYL frogs have undergone substantial evolutionary change
associated with Bd exposure and that might provide a genetic basis for
the increased resistance/tolerance to Bd infection observed in naturally
recovering populations \citep{knapp2016}. We then use the reintroduction
of resistant/tolerant frogs to 24 sites and results from long-term
capture-mark-recapture (CMR) surveys to describe the dynamics and
drivers of population establishment. These data are also used to
parameterize a population model that extends our inferences of
population establishment well beyond the temporal extent of the CMR
data. Collectively, these results provide one of the first examples of
how evolutionary rescue and the reintroduction of resistant/tolerant
animals can allow the large-scale recovery of disease-impacted species.

@@

To determine whether MYL frog populations show an evolutionary response
to Bd, we compared frog genomes between populations with contrasting
histories of Bd exposure. Specifically, we compared frog genomes sampled
in four populations that have not yet experienced a Bd-caused epizootic
(``naive'') \citep{zhou2015} versus in five populations that experienced
a Bd epizootic during the past several decades and have since recovered
to varying degrees (``recovering''; Figure~\ref{fig-allelefrequencies})
\citep{knapp2016, vredenburg2010}. These recovering populations exist in
an enzootic state \citep{briggs2010}, characterized by high Bd
prevalence and, in adults, moderate Bd infection intensities (``load'')
that are typically well below the level expected to cause mortality
\citep{vredenburg2010}.

We conducted a principal component analysis (PCA) of the genomic data to
describe the relationships between sampled populations, and then used
two complementary approaches to identify regions of the genome under
selection. First, we used a multivariate linear mixed model to evaluate
associations between population type (i.e., naive versus recovering) and
individual variants, including single nucleotide polymorphisms (SNPs)
and insertions/deletions (INDELS). Second, we used a splined window
analysis to identify larger genomic regions showing differences between
population types in \emph{F\textsubscript{ST}} and nucleotide diversity
(\(\pi_{diff}\) = \(\pi_{naive}\) - \(\pi_{recovering}\)).

Individual frogs clustered into 3 separate groups in principal component
space (Figure S1A), and clusters reflected the
strong signature of isolation by distance that is characteristic of MYL
frogs \citep{rothstein2020, poorten2017}. Importantly, each cluster
contained at least one population from both the naive and recovering
groups, allowing us to distinguish allelic associations of individuals
sampled in the 2 population types versus allelic associations resulting
from population structure and genetic drift.

Results from the individual variant and splined window analyses show
that recovering populations have signatures of strong selection on
multiple regions of the genome. The analysis of individual variants
identified 11 ``outlier'' SNPs (i.e., showing significantly different
allele frequencies between naive versus recovering populations) from 7
distinct genes across 4 contigs (Figure S1B,C). One of the 7 identified genes did not have an associated
annotation. Frequency differences between the naive and recovering
populations ranged from 41\% to 86\%. Most of these SNPs showed
frequency differences in only a subset of the sampled populations
(Figure~\ref{fig-allelefrequencies} A, B), but the SNP in the RIN3 gene
showed consistent differences in frequencies across all populations
(Figure~\ref{fig-allelefrequencies} C). This is suggestive of strong -
and parallel - selection at this locus associated with exposure of frogs
to Bd across multiple populations.

The splined window analysis identified 33 outlier regions for
\(\pi_{diff}\) and 58 outlier regions for \emph{F\textsubscript{ST}}
(Figure~\ref{fig-spline-manhattan} A, B). Of these, 9 regions were
outliers for both metrics and 2 of these shared regions also contained
outlier SNPs. A total of 33 annotated genes were found in the 9 shared
regions (SI). The largest \(\pi_{diff}\), indicative of directional
selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of
the RIN3 outlier SNP (Figure~\ref{fig-spline-manhattan} C). This region
contains approximately 500 SNPs and one annotated gene called
``interferon-induced very large GTPase 1-like'' (GVINP1). Additionally,
a shared outlier region on Contig1 contained two complement factor genes
(C6 and C7). Interestingly, this region had a low \(\pi_{diff}\) value,
consistent with balancing selection. Finally, one shared outlier region
on Contig8 contained one outlier SNP (TCF19) and was within 360kb of
another outlier SNP (VARS) (Figure~\ref{fig-spline-manhattan} D,
Figure S2). This region (854kb from the beginning of
the outlier window to the VARS SNP) contained a total of 8 annotated
genes. In \emph{Xenopus}, five of these genes occur in the extended
major histocompatibility complex (MHC) Class I region (FLOT1, TUBB,
MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III
region (HSP70, LSM2, VARS) \citep{ohta2006}. Therefore, this region
under selection is part of the extended MHC Class I and III complex and
shows synteny with other amphibian genomes.

@@

To determine whether Bd load had a negative effect on adult survival, as
would be expected if frogs were highly susceptible to Bd infection, we
developed linear and multilevel models. The best model of 1-year frog
survival identified several important predictors, but Bd load at the
time of translocation was not among them (bd\_load;
Figure S6). Important predictors included
winter severity in the year following translocation (snow\_t1), site
elevation (elevation), and donor population (donor\_72996;
Figure~\ref{fig-cond-effects}). Males had somewhat higher survival than
females, but this effect was only marginally important (sex\_male;
Figure~\ref{fig-cond-effects}, Figure S6).
Collectively, these results are consistent with frogs in recovering
populations having evolved sufficient resistance/tolerance to minimize
the adverse effects of Bd infection. In addition, the moderate Bd loads
that characterized frogs following translocation
(Figure S5) and relatively high survival of
frogs in numerous translocated cohorts
(Figure~\ref{fig-translocation-survival}) suggests that frogs
translocated from recovering populations can maintain the benefits of
resistance/tolerance in non-natal habitats. Finally, in 3 locations
where longer CMR time series allowed us to assess the survival of new
adults recruited to the population, naturally recruited adults had
equivalent or higher survival probabilities than the originally
translocated adults (Figure S7). Therefore,
frog resistance/tolerance in non-natal habitats is maintained across
generations.

@@

Even when incorporating demographic stochasticity and environmental
stochasticity in the probability of juvenile recruitment to the adult
class \(\omega\) (the transition that we expect to be the most subject
to environmental variability in the presence of Bd), populations with
high adult survival are likely to persist over a 50 year time horizon.
Our model predicted that, following a single introduction of 40 adult
individuals into a population and using the baseline value for
\(\omega\) of 0.3, the six populations with the highest adult survival
probabilities had 50-year extinction probabilities of 0 to 0.32;
Figure~\ref{fig-viability} B). This indicates strong potential for
long-term persistence in the presence of Bd and environmental
variability in recruitment (Figure~\ref{fig-viability} B). When we
decreased \(\omega\) from 0.3 to a conservative estimate of 0.15,
50-year extinction probability remained less than 0.5 for three
populations and was 0.79 to 0.94 for three additional populations
(Figure~\ref{fig-viability} B). In contrast, for the six populations
where yearly adult survival probability \(\sigma_{A_R} < 0.5\),
extinction probability over 50 years was always predicted to be 1. To
test the validity of our model predictions, we demonstrated that our
stochastic model could describe the general recovery trajectory of our
translocated population with the longest survey history
(Figure~\ref{fig-viability} C; population 70550, surveyed for 16 years).

@@

Results from our genomic analyses strongly suggest that natural
selection for adaptive alleles and resulting increased
resistance/tolerance is at least partially responsible for the reduced
susceptibility of frogs in recovering populations. We identified
specific alleles and genomic regions showing signatures of selection
between adjacent Bd-naive and recovering MYL frog populations,
consistent with strong selection following Bd exposure. Importantly,
some regions under selection are associated with cellular and
immunological mechanisms known to contribute to disease resistance,
including in amphibians \citep{zamudio2020}. For example, the MHC plays
an important role in immunity. In our study, we identified a region that
shows evidence of selection in recovering populations and contains eight
genes associated with either the MHC Class I or Class III regions. These
results corroborate numerous previous studies linking MHC genes to
amphibian resistance/tolerance against Bd
\citep[e.g.,][]{savage2011, bataille2015}. Similarly, the region with
the strongest indication of directional selection (as measured by
\(\pi_{diff}\)) contains the gene GVINP1. Several previous studies have
found this gene to be differentially expressed during Bd infection
\citep[e.g.,][]{grogan2018, ellison2014} and in populations differing in
Bd susceptibility \citep{grogan2018}. This gene is also strongly linked
to disease in salmon, explaining a notable 20\% of the resistance
phenotype \citep{robledo2020, robledo2018}. We also identified a region,
characterized by high \emph{F\textsubscript{ST}} and low \(\pi_{diff}\),
that contained the complement genes C6 and C7. This could indicate that
balancing selection is acting in this region of the genome to favor a
diverse set of alleles, as is known for C6 in humans
\citep{soejima2005}. Finally, the RIN3 gene showed a consistent pattern
of allele frequency differences across all nine of the frog populations
sampled in this study, indicating consistent selection in populations
distributed across a wide geographic area. This gene is associated with
inflammation response and in \emph{Xenopus} is highly expressed in wound
stumps following amputation of tadpole tails \citep{fukazawa2009}. In
the SI, we provide additional details on these and other genes
identified in our study as under selection.

@@

To visualize the genomic relationship of our populations we conducted a
principal components analysis (PCA) using the glPCA function in the
adegenet R package \citep{jombart2008}. To detect regions of the genome
under selection between naive and recovering populations, we used two
approaches: (1) a multivariate linear mixed model to evaluate individual
variants (SNPs and INDELs), and (2) a splined window analysis to
evaluate larger genomic regions. For the variant analysis, we first used
a stringent data filter to include only variants with \textless{} 5\%
missing data (missing for no more than 2 individuals), and then
calculated the likelihood ratio statistic for the resulting set of
148,307 high quality variants across 127 contigs using GEMMA. GEMMA
calculates and incorporates a relatedness matrix for input samples,
allowing us to account for relatedness and population structure when
calculating likelihood ratio statistics. We identified variants showing
different allele frequencies between naive versus recovering populations
(``outliers'') using a Bonferroni- corrected significance level of 0.01.
We visualized the results using a Manhattan plot and qqplot. We
developed a more liberal set of outlier variants using a Bonferroni-
corrected significance level of 0.05 and used this set solely for the
gene ontology (GO) analysis (see below and SI).

@@

We estimated survival of translocated frogs, recruitment of new frogs
into the adult population, and adult population size using a Bayesian
open-population Jolly-Seber mark-recapture model with known additions to
the population (i.e., translocated cohorts), as implemented by the mrmr
package \citep{joseph2019} and using R Statistical Software
\citep[v4.4.4,][]{rsoftware2022}. The model tracks the states of
\emph{M} individuals that comprise a superpopulation made up of real and
pseudo-individuals \citep[see][ for details]{joseph2018}. The possible
states of individuals include ``not recruited'', ``alive'', and
``dead''. The possible observations of individuals include ``detected''
and ``not detected''. We assume that individuals that are in the ``not
recruited'' or ``dead'' states are never detected (i.e., there are no
mistakes in the individual PIT tag records). For all models, we used
mrmr defaults for priors, number of chains (4), and warmup and
post-warmup iterations (2000 for each). We evaluated convergence of the
Markov chain Monte Carlo (MCMC) algorithm using trace plots and
Gelman-Rubin statistics (Rhat). Code used to create CMR datasets and fit
models for each translocated population is available from the following
GitHub repository: \url{https://github.com/SNARL1/cmr-analysis}.

@@

After parameterizing our model with CMR-estimated adult frog survival
probabilities and other known vital rates (Table S2), we performed four analyses (described in detail in the SI).
First, we compute the long-run growth rate \(\lambda\) for each of our
12 translocated populations to determine if the populations were
deterministically predicted to grow or decline in the long-run. Second,
we compute the elasticity of \(\lambda\) to model parameters, to
quantify how much changes in adult survival, for example, affected the
long-run growth rate. Third, we included demographic and environmental
stochasticity into our model and simulated the 50-year viability (i.e.,
1 - extinction probability) of populations given an introduction of 40
adult individuals into an unoccupied habitat. Finally, we fit our model
to our longest translocation trajectory to confirm that our model could
reasonably reproduce the observed recovery trajectories of MYL frogs
following reintroductions.}

@@

{\centering \includegraphics[width=\textwidth]{figures/allele_maps.png}

@@

{\centering \includegraphics[width=\textwidth]{figures/splinewindow_manhattan.png}

@@

{\centering \includegraphics[width=\textwidth]{figures/translocation_survival_bysiteid.png}

@@

{\centering \includegraphics[width=\textwidth]{figures/cond_effects_plot.png}
